Analysis: Machine Learning - ReqAmt

### 2021-11-28 code/analysis_code/analysis-ML-RA-outlier.Rmd
Outlier Analysis

---
title: "Analysis: Machine Learning - OblAmt - Outliers"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---

@@

This script uses obligated FEMA funds as the outcome of interest and fits the following models to the analysis data:

@@

It compares the  models, and then finally fits the “best” model to the test data. It removes the $4B outlier that corresponds to Hurricane Maria in Puerto Rico in 2017.

@@

* Drop the unused variables
* Perform feature engineering
* Transform Incident Month into ordinal variable
* Remove outlier
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on incidentType
* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
obl_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, OblAmt)

#transform incident month into ordinal variable
obl_data$IncidentMonth <- ordered(obl_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

@@

obl_data$FEMARegion <- ifelse((obl_data$state == "CT" |
                                 obl_data$state == "ME" |
                                 obl_data$state == "MA" |
                                 obl_data$state == "NH" |
                                 obl_data$state == "RI" |
                                 obl_data$state == "VT"),

@@

                              ifelse((obl_data$state == "NJ" |
                                         obl_data$state == "NY" |
                                         obl_data$state == "PR" |
                                         obl_data$state == "VI"),

@@

                                       ifelse((obl_data$state == "DE" |
                                                 obl_data$state == "MD" |
                                                 obl_data$state == "PA" |
                                                 obl_data$state == "VA" |
                                                 obl_data$state == "DC" |
                                                 obl_data$state == "WV"),

@@

                                               ifelse((obl_data$state == "AL" |
                                                         obl_data$state == "FL" |
                                                         obl_data$state == "GA" |
                                                         obl_data$state == "KY" |
                                                         obl_data$state == "MS" |
                                                         obl_data$state == "NC" |
                                                         obl_data$state == "SC" |
                                                         obl_data$state == "TN"),

@@

                                                      ifelse((obl_data$state == "IL" |
                                                               obl_data$state == "IN" |
                                                               obl_data$state == "MI" |
                                                               obl_data$state == "MN" |
                                                               obl_data$state == "OH" |
                                                               obl_data$state == "WI"),

@@

                                                             ifelse((obl_data$state == "AR" |
                                                                         obl_data$state == "LA" |
                                                                         obl_data$state == "NM" |
                                                                         obl_data$state == "OK" |
                                                                         obl_data$state == "TX"),

@@

                                                                       ifelse((obl_data$state == "IA" |
                                                                                 obl_data$state == "KS" |
                                                                                 obl_data$state == "MO" |
                                                                                 obl_data$state == "NE"),

@@

                                                                               ifelse((obl_data$state == "CO" |
                                                                                         obl_data$state == "MT" |
                                                                                         obl_data$state == "ND" |
                                                                                         obl_data$state == "SD" |
                                                                                         obl_data$state == "UT" |
                                                                                         obl_data$state == "WY"),

@@

                                                                                       ifelse((obl_data$state == "AZ" |
                                                                                                 obl_data$state == "CA" |
                                                                                                 obl_data$state == "HI" |
                                                                                                 obl_data$state == "NV"),

@@

                                                                                               ifelse((obl_data$state == "AK" |
                                                                                                         obl_data$state == "ID" |
                                                                                                         obl_data$state == "OR" |
                                                                                                         obl_data$state == "WA"),

@@

#drop state variable
obl_data <- obl_data %>%
              dplyr::select(-c(state))

#add to "other" category for incident Types that are infrequent
obl_data$incidentType[(obl_data$incidentType == "Chemical" |
                        obl_data$incidentType == "Coastal Storm" |
                        obl_data$incidentType == "Dam/Levee Break" |
                        obl_data$incidentType == "Mud/Landslide" |
                        obl_data$incidentType == "Volcano" |
                        obl_data$incidentType == "Terrorist" |
                        obl_data$incidentType == "Snow")] <- "Other"

#remove Hurricane Maria outlier
obl_data <-subset(obl_data, OblAmt< 4000000000)

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
obl_data_split <- rsample::initial_split(obl_data, prop = 2/3,
                                     strata = incidentType)

#create dataframes for the two sets:
obl_train_data <- rsample::training(obl_data_split)
obl_test_data <- rsample::testing(obl_data_split)

#5-fold cross validation, repeated 5 times, stratified by incidentType
obl_folds <- rsample::vfold_cv(obl_train_data,
                               strata = incidentType,
                               v = 5,
                               repeats = 5)

#create recipe that codes categorical variables as dummy variables
OA_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#add recipe and model into workflow
null_wflow <- workflows::workflow() %>%
              workflows::add_recipe(OA_rec) %>%
              workflows::add_model(null_mod)

#"fit" model to training data
obl_null_train <- null_wflow %>%
                  parsnip::fit(data = obl_train_data)

#summary of null model with training data to get mean (which in this case is the RMSE)
obl_null_train_sum <- broom.mixed::tidy(obl_null_train)
obl_null_train_sum

#RMSE for training data for formatting later
obl_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_train_data$OblAmt,
                                              estimate = rep(mean(obl_train_data$OblAmt), nrow(obl_train_data))),
                          SE = 0,
                          model = "Null - Train")

#"fit" model to test data
obl_null_test <- null_wflow %>%
                  parsnip::fit(data = obl_test_data)

#summary of null model with test data to get mean (which in this case is the RMSE)
obl_null_test_sum <- broom.mixed::tidy(obl_null_test)
obl_null_test_sum

#RMSE for testing data for formatting later
obl_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = obl_test_data$OblAmt,
                                              estimate = rep(mean(obl_test_data$OblAmt), nrow(obl_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#use the recipe specified earlier
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
tree_wflow <- workflows::workflow() %>%
               workflows::add_model(tree_mod) %>%
               workflows::add_recipe(OA_rec)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = obl_folds,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot-obl.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

@@

#save file
DT_detail_fig_file = here("results","DT-det-plot-obl.png")
ggsave(filename = DT_detail_fig_file, plot = DT_detail_plot)

@@

#one last fit on the training data
best_tree_fit <- best_tree_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                         roundint = F,
                                         type = 5,
                                         digits = 5,
                                         main = "Selected Decision Tree Model")
#save file manually -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
DT_pred_act

#save file
DT_pred_act_file = here("results","DT-pred-act-obl.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

@@

#save file
DT_resid_file = here("results","DT-resid-obl.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
DT_pred_resid

#save file
DT_pred_resid_file = here("results","DT-pred-resid-obl.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

#compare model performance to null model
tree_RMSE <- tree_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Tree") %>%
               dplyr::bind_rows(obl_null_RMSE_train)
tree_RMSE
```

@@

### 2. Workflow Definition
```{r}
#define workflow for RF regression
RF_wflow <- workflows::workflow() %>%
               workflows::add_model(RF_mod) %>%
               workflows::add_recipe(OA_rec)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = obl_folds,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot-obl.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#one last fit on the training data
best_RF_fit <- best_RF_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

@@

#save  manually - ggsave throws an error

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = obl_train_data) %>%
                dplyr::select(.pred, OblAmt) %>%
                dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point() +
                        labs(title = "RF Fit: Actual vs. Predicted Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")

#save file
RF_pred_act_file = here("results","RF-pred-act-obl.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

@@

#save file
RF_resid_file = here("results","RF-resid-obl.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Obligated Funds",
                                x = "Fitted Requested Funds ($)",
                                y = "Residual")
RF_pred_resid

#save file
RF_pred_resid_file = here("results","RF-pred-resid-obl.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

@@

#use bootstrap instead of CV
obl_data_bootsplit <- obl_train_data %>%
                        rsample::bootstraps(times = 10,
                                            strata = "incidentType")

@@

#create recipe that codes categorical variables as dummy variables
BT_rec <- recipes::recipe(OblAmt ~ ., data = obl_train_data) %>%
          step_string2factor(all_nominal())
```
---

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
BT_res <- BT_wflow %>%
          tune::tune_grid(resamples = obl_data_bootsplit,
                          grid = tree_grid,
                          metrics = metric_set(rmse),
                          control = control_resamples(save_pred = TRUE,
                                                      extract = function(x) extract_model(x)))

@@

#save file
BT_auto_fig_file = here("results","BT-auto-plot-obl.png")
ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)

@@

#save file
BT_det_fig_file = here("results","BT-det-plot-obl.png")
ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
```

@@

#one last fit on the training data
best_BT_fit <- best_BT_wflow %>%
                    parsnip::fit(data = obl_train_data)
```

@@

### 6. Model evaluation
```{r}
#extract model from final fit
x_BT <- best_BT_fit$fit$fit$fit

#bag roots function to extract fit
bag_roots <-  function(x){
  x %>% 
    dplyr::select(.extracts) %>% 
    unnest(cols = c(.extracts)) %>% 
    dplyr::mutate(models = map(.extracts,
                               ~.x$model_df)) %>% 
    dplyr::select(-.extracts) %>% 
    unnest(cols = c(models)) %>% 
    mutate(root = map_chr(model,
                          ~as.character(.x$fit$frame[1, 1]))) %>%
    dplyr::select(root)  
}

@@

#find predictions and intervals
BT_resid <- best_BT_fit %>%
                  broom.mixed::augment(new_data = obl_train_data) %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
BT_pred_act

#save file
BT_pred_act_file = here("results","BT-pred-act-obl.png")
ggsave(filename = BT_pred_act_file, plot = BT_pred_act)

@@

#save file
BT_resid_file = here("results","BT-resids-obl.png")
ggsave(filename = BT_resid_file, plot = BT_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
BT_pred_resid

#save file
BT_pred_resid_file = here("results","BT-pred-resid-obl.png")
ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)

@@

#save combined table
model_comp_file = here("results", "ML-model-comp-obl.Rds")
saveRDS(BT_RMSE, file = model_comp_file)
```

@@

## Model Selection and Evaluation
Using the bagged tree model
```{r}
#fit model to training set but evaluate on the test data
BT_fit_test <- best_BT_wflow %>%
                    tune::last_fit(split = obl_data_split)

@@

#find predictions and intervals
BT_resid_fit <- BT_fit_test %>%
                  broom.mixed::augment() %>%
                  dplyr::select(.pred, OblAmt) %>%
                  dplyr::mutate(.resid = .pred - OblAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = OblAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Obligated Funds",
                              x = "Predicted Obligated Funds ($)",
                              y = "Actual Obligated Funds ($)")
BT_pred_act_test

#save file
BT_pred_act_test_file = here("results","BT-pred-act-test-obl.png")
ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)

@@

#save file
BT_resid_test_file = here("results","BT-resid-test-obl.png")
ggsave(filename = BT_resid_test_file, plot = BT_resids_test)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Obligated Funds",
                                x = "Predicted Obligated Funds ($)",
                                y = "Residual")
BT_pred_resid_test

>>>>

---
title: "Analysis: Machine Learning - ReqAmt"
output: 
  html_document:
    theme: flatly
    toc: FALSE
---

@@

This script uses requested FEMA funds as the outcome of interest and fits the following models to the analysis data:

@@

It compares the  models, and then finally fits the “best” model to the test data.

@@

* Drop the unused variables
* Perform feature engineering
* Transform Incident Month into ordinal variable
* Remove outlier
* Set the random seed to 123
* Split the dataset into 70% training, 30% testing, stratified on incidentType
* 5-fold cross validation, 5 times repeated, stratified on incidentType for the CV folds
* Create a recipe for data and fitting that codes categorical variables as dummy variables
```{r data setup}
#requested amount filter
req_data <- analysisdata %>%
              dplyr::select(state, IncidentYear, IncidentMonth, declarationType, incidentType, IncidentDuration, ResponseDays, IH, PA, HM,
                            Population, Counties, TotalAgencies, FEMACSAvg, ReqAmt)

#transform incident month into ordinal variable
req_data$IncidentMonth <- ordered(req_data$IncidentMonth, levels = 1:12,
                                  labels = c("January", "February", "March", "April", "May", "June", "July", "August", "September", "October", "November", "December"))

@@

req_data$FEMARegion <- ifelse((req_data$state == "CT" |
                                 req_data$state == "ME" |
                                 req_data$state == "MA" |
                                 req_data$state == "NH" |
                                 req_data$state == "RI" |
                                 req_data$state == "VT"),

@@

                              ifelse((req_data$state == "NJ" |
                                         req_data$state == "NY" |
                                         req_data$state == "PR" |
                                         req_data$state == "VI"),

@@

                                       ifelse((req_data$state == "DE" |
                                                 req_data$state == "MD" |
                                                 req_data$state == "PA" |
                                                 req_data$state == "VA" |
                                                 req_data$state == "DC" |
                                                 req_data$state == "WV"),

@@

                                               ifelse((req_data$state == "AL" |
                                                         req_data$state == "FL" |
                                                         req_data$state == "GA" |
                                                         req_data$state == "KY" |
                                                         req_data$state == "MS" |
                                                         req_data$state == "NC" |
                                                         req_data$state == "SC" |
                                                         req_data$state == "TN"),

@@

                                                      ifelse((req_data$state == "IL" |
                                                               req_data$state == "IN" |
                                                               req_data$state == "MI" |
                                                               req_data$state == "MN" |
                                                               req_data$state == "OH" |
                                                               req_data$state == "WI"),

@@

                                                             ifelse((req_data$state == "AR" |
                                                                         req_data$state == "LA" |
                                                                         req_data$state == "NM" |
                                                                         req_data$state == "OK" |
                                                                         req_data$state == "TX"),

@@

                                                                       ifelse((req_data$state == "IA" |
                                                                                 req_data$state == "KS" |
                                                                                 req_data$state == "MO" |
                                                                                 req_data$state == "NE"),

@@

                                                                               ifelse((req_data$state == "CO" |
                                                                                         req_data$state == "MT" |
                                                                                         req_data$state == "ND" |
                                                                                         req_data$state == "SD" |
                                                                                         req_data$state == "UT" |
                                                                                         req_data$state == "WY"),

@@

                                                                                       ifelse((req_data$state == "AZ" |
                                                                                                 req_data$state == "CA" |
                                                                                                 req_data$state == "HI" |
                                                                                                 req_data$state == "NV"),

@@

                                                                                               ifelse((req_data$state == "AK" |
                                                                                                         req_data$state == "ID" |
                                                                                                         req_data$state == "OR" |
                                                                                                         req_data$state == "WA"),

@@

#drop state variable
req_data <- req_data %>%
              dplyr::select(-c(state))

#add to "other" category for incident Types that are infrequent
req_data$incidentType[(req_data$incidentType == "Chemical" |
                        req_data$incidentType == "Coastal Storm" |
                        req_data$incidentType == "Dam/Levee Break" |
                        req_data$incidentType == "Mud/Landslide" |
                        req_data$incidentType == "Volcano" |
                        req_data$incidentType == "Terrorist" |
                        req_data$incidentType == "Snow")] <- "Other"

#remove Hurricane Maria outlier
req_data <-subset(req_data, ReqAmt< 4000000000)

@@

#split dataset into 66.6% training, 33.3% testing
#use incidentType as stratification
req_data_split <- rsample::initial_split(req_data, prop = 2/3,
                                     strata = incidentType)

#create dataframes for the two sets:
req_train_data <- rsample::training(req_data_split)
req_test_data <- rsample::testing(req_data_split)

#5-fold cross validation, repeated 5 times, stratified by incidentType
req_folds <- rsample::vfold_cv(req_train_data,
                               strata = incidentType,
                               v = 5,
                               repeats = 5)

#create recipe that codes categorical variables as dummy variables
RA_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
          recipes::step_dummy(all_nominal_predictors())
```

@@

#add recipe and model into workflow
null_wflow <- workflows::workflow() %>%
              workflows::add_recipe(RA_rec) %>%
              workflows::add_model(null_mod)

#"fit" model to training data
req_null_train <- null_wflow %>%
                  parsnip::fit(data = req_train_data)

#summary of null model with training data to get mean (which in this case is the RMSE)
req_null_train_sum <- broom.mixed::tidy(req_null_train)
req_null_train_sum

#RMSE for training data for formatting later
req_null_RMSE_train <- tibble::tibble(
                          rmse = rmse_vec(truth = req_train_data$ReqAmt,
                                              estimate = rep(mean(req_train_data$ReqAmt), nrow(req_train_data))),
                          SE = 0,
                          model = "Null - Train")

#"fit" model to test data
req_null_test <- null_wflow %>%
                  parsnip::fit(data = req_test_data)

#summary of null model with test data to get mean (which in this case is the RMSE)
req_null_test_sum <- broom.mixed::tidy(req_null_test)
req_null_test_sum

#RMSE for testing data for formatting later
req_null_RMSE_test <- tibble::tibble(
                          rmse = rmse_vec(truth = req_test_data$ReqAmt,
                                              estimate = rep(mean(req_test_data$ReqAmt), nrow(req_test_data))),
                          SE = 0,
                          model = "Null - Test")
```

@@

#use the recipe specified earlier (line 133)
```
---

### 2. Workflow Definition
```{r}
#define workflow for tree
tree_wflow <- workflows::workflow() %>%
               workflows::add_model(tree_mod) %>%
               workflows::add_recipe(RA_rec)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
tree_res <- tree_wflow %>%
                tune::tune_grid(resamples = req_folds,
                                grid = tree_grid,
                                control = control_grid(verbose = TRUE),
                                metrics = yardstick::metric_set(rmse))

@@

#save file
DT_auto_fig_file = here("results","DT-auto-plot.png")
ggsave(filename = DT_auto_fig_file, plot = DT_auto_plot)

@@

#save file
DT_det_fig_file = here("results","DT-det-plot.png")
ggsave(filename = DT_det_fig_file, plot = DT_detail_plot)

@@

#one last fit on the training data
best_tree_fit <- best_tree_wflow %>%
                    parsnip::fit(data = req_train_data)
```

@@

### 6. Model evaluation
```{r}
#plot the tree
rpart.plot::rpart.plot(x = workflowsets::extract_fit_parsnip(best_tree_fit)$fit,
                                                     roundint = F,
                                                     type = 5,
                                                     digits = 5,
                                                     main = "Selected Decision Tree Model")
#save manually -- ggsave throws an error

#find predictions and intervals
tree_resid <- best_tree_fit %>%
                  broom.mixed::augment(new_data = req_train_data) %>%
                  dplyr::select(.pred, ReqAmt) %>%
                  dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
DT_pred_act <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Decision Tree Fit: Predicted vs. Actual Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
DT_pred_act

#save file
DT_pred_act_file = here("results","DT-pred-act.png")
ggsave(filename = DT_pred_act_file, plot = DT_pred_act)

@@

#save file
DT_resid_file = here("results","DT-resid.png")
ggsave(filename = DT_resid_file, plot = DT_resid)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
DT_pred_resid <- ggplot2::ggplot(tree_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Decision Tree Fit: Residuals vs Fitted Requested Funds",
                                x = "Predicted Requested Funds ($)",
                                y = "Residual")
DT_pred_resid

#save file
DT_pred_resid_file = here("results","DT-pred-resid.png")
ggsave(filename = DT_pred_resid_file, plot = DT_pred_resid)

@@

#compare model performance to null model
tree_RMSE <- tree_res %>%
                tune::show_best(n = 1) %>%
                dplyr::transmute(
                  rmse = round(mean, 3),
                  SE = round(std_err, 4),
                  model = "Tree") %>%
               dplyr::bind_rows(req_null_RMSE_train)
tree_RMSE
```

The identified best performing tree model has some interesting results. It identifies incident duration, state population, total number of federal agencies, total number of state counties, and individual/households program awardance as significant decision points for predicting requesting FEMA funding. All of these variables are to be expected, but it is interesting that incident type or declaration type are not included as significant. The RMSE is improved from the Null model, but there's still opportunity for improvement. Let's try another model.

@@

### 2. Workflow Definition
```{r}
#define workflow for RF regression
RF_wflow <- workflows::workflow() %>%
               workflows::add_model(RF_mod) %>%
               workflows::add_recipe(RA_rec)
```

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
RF_res <- RF_wflow %>%
              tune::tune_grid(resamples = req_folds,
                              grid = RF_grid,
                              control = control_grid(verbose = TRUE, save_pred = TRUE),
                              metrics = metric_set(rmse))

@@

#save file
RF_auto_fig_file = here("results","RF-auto-plot.png")
ggsave(filename = RF_auto_fig_file, plot = RF_auto_plot)
```

@@

#one last fit on the training data
best_RF_fit <- best_RF_wflow %>%
                    parsnip::fit(data = req_train_data)
```

@@

#save manually-- ggsave throws errors

#find predictions and intervals
RF_resid <- best_RF_fit %>%
                broom.mixed::augment(new_data = req_train_data) %>%
                dplyr::select(.pred, ReqAmt) %>%
                dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline is a 45 degree line, along which the results should fall
RF_pred_act <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) +
                        geom_point() +
                        labs(title = "RF Fit: Actual vs. Predicted Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
RF_pred_act

#save file
RF_pred_act_file = here("results","RF-pred-act.png")
ggsave(filename = RF_pred_act_file, plot = RF_pred_act)

@@

#save file
RF_resid_file = here("results","RF-resid.png")
ggsave(filename = RF_resid_file, plot = RF_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
RF_pred_resid <- ggplot2::ggplot(RF_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Random Forest Fit: Residuals vs Fitted Requested Funds",
                                x = "Fitted Requested Funds ($)",
                                y = "Residual")
RF_pred_resid

#save file
RF_pred_resid_file = here("results","RF-pred-resid.png")
ggsave(filename = RF_pred_resid_file, plot = RF_pred_resid)

@@

In examining the results of the RF model within the context of RMSE, it performs marginally better than the decision tree. Interestingly, it includes declaration month of Jan - March as more important than state population. Let's try one more model.

@@

#use bootstrap instead of CV
req_data_bootsplit <- req_train_data %>%
                        rsample::bootstraps(times = 10,
                                            strata = "incidentType")

@@

#create recipe that codes categorical variables as dummy variables
BT_rec <- recipes::recipe(ReqAmt ~ ., data = req_train_data) %>%
          step_string2factor(all_nominal())
```
---

@@

### 4. Tuning Using Cross-Validation and the `tune_grid()` function
```{r}
#tune the model with previously specified cross-validation and RMSE as target metric
BT_res <- BT_wflow %>%
          tune::tune_grid(resamples = req_data_bootsplit,
                          grid = tree_grid,
                          metrics = metric_set(rmse),
                          control = control_resamples(save_pred = TRUE,
                                                      extract = function(x) extract_model(x)))

@@

#save file
BT_auto_fig_file = here("results","BT-auto-plot.png")
ggsave(filename = BT_auto_fig_file, plot = BT_auto_plot)

@@

#save file
BT_det_fig_file = here("results","BT-det-plot.png")
ggsave(filename = BT_det_fig_file, plot = BT_det_plot)
```

@@

#one last fit on the training data
best_BT_fit <- best_BT_wflow %>%
                    parsnip::fit(data = req_train_data)
```

@@

### 6. Model evaluation
```{r}
#bag roots function to extract most important variables
bag_roots <-  function(x){
  x %>% 
    dplyr::select(.extracts) %>% 
    unnest(cols = c(.extracts)) %>% 
    dplyr::mutate(models = map(.extracts,
                               ~.x$model_df)) %>% 
    dplyr::select(-.extracts) %>% 
    unnest(cols = c(models)) %>% 
    mutate(root = map_chr(model,
                          ~as.character(.x$fit$frame[1, 1]))) %>%
    dplyr::select(root)  
}

@@

#find predictions and intervals
BT_resid <- best_BT_fit %>%
                  broom.mixed::augment(new_data = req_train_data) %>%
                  dplyr::select(.pred, ReqAmt) %>%
                  dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Fit: Predicted vs. Actual Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
BT_pred_act

#save file
BT_pred_act_file = here("results","BT-pred-act.png")
ggsave(filename = BT_pred_act_file, plot = BT_pred_act)

@@

#save file
BT_resid_file = here("results","BT-resid.png")
ggsave(filename = BT_resid_file, plot = BT_resids)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid <- ggplot2::ggplot(BT_resid, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Fit: Residuals vs Fitted Requested Funds",
                                x = "Predicted Requested Funds ($)",
                                y = "Residual")
BT_pred_resid

#save file
BT_pred_resid_file = here("results","BT-pred-resid.png")
ggsave(filename = BT_pred_resid_file, plot = BT_pred_resid)

@@

#save combined table
model_comp_file = here("results", "ML-model-comp.Rds")
saveRDS(BT_RMSE, file = model_comp_file)
```

@@

## Model Selection and Evaluation
Using the bagged tree model
```{r}
#fit model to training set but evaluate on the test data
BT_fit_test <- best_BT_wflow %>%
                    tune::last_fit(split = req_data_split)

@@

#find predictions and intervals
BT_resid_fit <- BT_fit_test %>%
                  broom.mixed::augment() %>%
                  dplyr::select(.pred, ReqAmt) %>%
                  dplyr::mutate(.resid = .pred - ReqAmt)

#plot model predictions from tuned model versus actual outcomes
#geom_abline draws a 45 degree line, along which the results should fall
BT_pred_act_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = ReqAmt)) +
                        geom_abline(slope = 1, intercept = 0, color = "red", lty = 2) + 
                        geom_point() +
                        labs(title = "Bagged Tree Test Fit: Predicted vs. Actual Requested Funds",
                              x = "Predicted Requested Funds ($)",
                              y = "Actual Requested Funds ($)")
BT_pred_act_test

#save file
BT_pred_act_test_file = here("results","BT-pred-act-test.png")
ggsave(filename = BT_pred_act_test_file, plot = BT_pred_act_test)

@@

#save file
BT_resid_test_file = here("results","BT-resid-test.png")
ggsave(filename = BT_resid_test_file, plot = BT_resids_test)

#plot model fit vs residuals
#the geom_hline plots a straight horizontal line along which the results fall
BT_pred_resid_test <- ggplot2::ggplot(BT_resid_fit, aes(x = .pred, y = .resid))+
                          geom_hline(yintercept = 0, color = "red", lty = 2) +
                          geom_point() +
                          labs(title = "Bagged Tree Test Fit: Residuals vs Fitted Requested Funds",
                                x = "Predicted Requested Funds ($)",
                                y = "Residual")
BT_pred_resid_test

#save file
BT_pred_resid_test_file = here("results","BT-pred-resid-test.png")
ggsave(filename = BT_pred_resid_test_file, plot = BT_pred_resid_test)
```
